\chapter*{Acknowledgment}
I would like to extend my sincere gratitude to Professor Patrick Bours who has been my supervisor for three projects, including this thesis. Your passion for our field of study is highly motivating, and I cannot begin to describe how important your guidance and encouragement has been to me. Even in your most hectic workdays, you have always been available and found time to help me focus on what mattered the most. Thank you.

To my family, I think you know how much you mean to me. You have always encouraged me throughout the five years I have spent in Gjøvik. The time I have spent with you during my trips back home has been invaluable. Thank you. To my parents, Anita and Per-Eirik, thank you for your support and patience especially during the last 8 months.

To Sigve, who tempted me into coming along to Gjøvik, thank you for the years we spent together as roommates and for dragging me along. You are truly awesome. To my other friends back home, there are un(?)fortunately too many of you to mention everyone by name. Still, you know who you are. Thanks for being great. I do however want to mention Pål, who has given me advice and boosts of confidence throughout my education. Thank you.

To the guys in my study group, that being Audun, Emil, Erik, Jardar, Jørgen and Kristian, thanks for making these last couple of years of study so much more enjoyable. Studying for all our exams would have been a huge pain in the neck without your presence. Lastly, I would like to thank my friends in Klatregruppa Grad 9. Climbing and hanging out with you guys has been a ton of fun. 

\begin{flushright}
P-K.N.\\[1pc]
\end{flushright}